---
title: "Analysis of COVID-19 Testing in India"
output: 
     html_document: 
     highlight: pygments
     theme: spacelab
---

```{r load-packages, echo=FALSE, message = FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)
library(statsr)
library(gridExtra)
library(janitor)
```
     
## Introduction

Till mid-April, India was handling the COVID-19 situation well with its 1.25 billion population in comparison to all the other countries(such as US and other countries) around the world.

But the situation changed pretty quickly at the end of June 2020. India is now among the top 5 countries with the highest number of cases.

We all have heard multiple times from many health care representatives that the only way to contain the spread of COVID-19 is *testing, testing, and testing*. Therefore, it is important to first analyze the impact of COVID-19 testing done in containing the spread in different countries. Using learnings from that analysis, we can understand how badly or well India and different Indian states are performing.

## Data source and tools

For this project, the COVID-19 testing data and its results in different countries were
obtained from [Our World in Data](https://ourworldindata.org/coronavirus-testing#source-information-country-by-country). The data on testing in different Indian states, was
found from [covid19india.org](https://api.covid19india.org/) 

All the analysis has been done using RStudio(R - version 4.0) and the output HTML file has been generated using R-Mardown. (most snippets of code have been hidden to reduce the length of file).

```{r load-data, echo=FALSE}
world_covid_data <- read.csv("owid-covid-data.csv", sep = ",")
india_covid_data <- read.csv("statewise_tested_numbers_data.csv", sep = ",")

```

```{r data_cleanup_world, echo=FALSE}
world_covid_data <- world_covid_data %>%
                        mutate(days = round(as.numeric(difftime(date, "2019-12-31", unit="days"))))
world_covid_data <- world_covid_data %>%
                        mutate(new_testing = total_tests - lag(total_tests))

countries <- c("India", "United States", "Canada", "Russia")

filtered_world_covid <- world_covid_data %>%
                                filter(location %in% countries) %>%
                                filter(new_cases != 0, new_testing != 0) %>%
                                mutate(test_per_positive = new_testing / new_cases) %>%
                                mutate(death_rate = (new_deaths / new_cases) * 100)
```


## Testing Analysis - India vs World

The covid-19 statistics of United States, Russia, and Canada have been compared angainst India's statistics. Even though United States and Russia have high number of total covid-19 cases(like India), these countries have been chosen because they are in completely different stages of spread of coronavirus when compared to India. Canada has been chosen to represent a country which has performed well in containment of the virus.

First, let's have a look the number of daily cases across these nations.
```{r, echo=FALSE, message=FALSE,warning=FALSE}
filtered_world_covid %>% 
        ggplot() +
        geom_line(aes(x = days, y = new_cases, color = location)) +
        geom_smooth(aes(x = days, y = new_cases, color = location), size = 0.5) +
        geom_text(aes(x = 150, y = 30000, color = "United States", label = "United States")) +
        geom_text(aes(x = 160, y = 15000, color = "India", label = "India"), family = "sans") +
        geom_text(aes(x = 120, y = 12000, color = "Russia", label = "Russia")) +
        geom_text(aes(x = 160, y = 3000, color = "Canada", label = "Canada")) +
        ylim(0, 50000) +
        theme_classic()+
        theme(legend.position = "none") +
        labs(x = "Days", 
             y = "Number of cases",
             title = "Number of daily COVID-19 cases in different countries",
             caption = "Data source: Our World in Data") +
        theme(plot.title = element_text(size = 12),
              text = element_text(family = "sans", color = "gray25"),
              plot.caption = element_text(color = "gray30"),
              plot.background = element_rect(fill = "gray95"),
              plot.margin = unit(c(5, 10, 5, 10), units = "mm"))
```

Please note: Day 0 refers to 31^th^ December, 2019

Unlike other 3 countries, India's exponential growth of coronavirus cases clearly stands out. Russia has clearly moved out it's exponential growth. Canda seems to be performing well from the start.

Looking at this graph, one might conclude that India is not doing enough testing. So, to verify this conclusion, let's compare the daily testing of each of these nations.

```{r, echo=FALSE}
filtered_world_covid %>% 
        ggplot() +
        geom_line(aes(x = days, y = new_testing/1000, color = location)) +
        geom_text(aes(x = 120, y = 400, color = "United States", label = "United States")) +
        geom_text(aes(x = 150, y = 150, color = "India", label = "India"), family = "sans") +
        geom_text(aes(x = 160, y = 380, color = "Russia", label = "Russia")) +
        geom_text(aes(x = 170, y = 80, color = "Canada", label = "Canada")) +
        theme_classic()+
        theme(legend.position = "none") +
        labs(x = "Days", 
             y = "Number of tests (in thousands)",
             title = "Number of daily COVID-19 tests performed in different countries",
             caption = "Data source: Our World in Data") +
        theme(plot.title = element_text(size = 12),
              text = element_text(family = "sans", color = "gray25"),
              plot.caption = element_text(color = "gray30"),
              plot.background = element_rect(fill = "gray95"),
              plot.margin = unit(c(5, 10, 5, 10), units = "mm"))
```


Surprisingly, India has started performing 2 Lakhs(or 200K) per day and it is ramping up it's testing linearly like all the countries. Having said that, India, as expected previously, is not performing enough tests to control the spread.

The next logical question is *how much testing is enough?*

## Tests per positive
Tests per positive case refers to the number of tests performed per positive case$(Test\ per\ positive = Number\ of\ tests/ Number\ of\ positive\ cases)$. WHO has recommended that countries should perform at least [10 to 30 tests per positive case discovered](https://ourworldindata.org/coronavirus-testing#note-2)

Let's see how these nations perform with respect to tests per positive.
```{r, echo=FALSE}
filtered_world_covid <- filtered_world_covid %>%
                                mutate(testing_level = case_when(
                                        between(test_per_positive, 0, 15) ~ "Below WHO recommendation",
                                        between(test_per_positive, 16, 30) ~ "On-par with WHO recommendation",
                                        T ~ "Better than WHO recommendation"
                                )) %>%
                                mutate(testing_level = factor(testing_level, 
                                                              levels = c("Below WHO recommendation", 
                                                                         "On-par with WHO recommendation", 
                                                                         "Better than WHO recommendation")))
```

```{r plot, echo=FALSE}
filtered_world_covid %>% 
        ggplot() +
        geom_col(aes(x = days, y = new_cases, fill = testing_level)) +
        labs(y = "New Cases", x = "Days") +
        facet_wrap(~ location, scales = "free") +
        labs(x = "Days", 
             y = "Number of cases",
             title = "Number of daily COVID-19 cases in different countries",
             caption = "Data source: Our World in Data") +
        theme(plot.title = element_text(size = 12),
              text = element_text(family = "sans", color = "gray25"),
              plot.caption = element_text(color = "gray30"),
              plot.background = element_rect(fill = "gray95"))+
        scale_fill_manual(values = c("red", "#FFE200", "#00C700"),
                          name = "Testing Level",
                          labels = c("Below WHO \nrecommendation(<15)",
                                     "On-par with WHO\nrecommendation(15-30)",
                                     "Better than WHO\nrecommendation(>30)"))
```

It's easy to figure that going into "Yellow" or "Green" region has helped the 3 countries in reducing the increase in number of new cases. In other words, *testing more than 15 people per positive case has helped US, Canada and Russia in slowing down the number of cases*. 

One can easily argue that reduction in increase of number of cases with increased testing is purely co-incidental for these 3 countries. 
Let's check correlation using Kendall's method between tests per positive and the number of new cases. Since many countries were not prepared for COVID-19 untill March 2020, the data from Jan 1, 2020 to Mar 31, 2020 has been removed.
```{r, message=FALSE, warning=FALSE, echo=FALSE}
correlation_test <- world_covid_data %>%
                                filter(days > 90) %>%
                                filter(new_cases > 5, new_testing != 0) %>%
                                mutate(test_per_positive = new_testing / new_cases) %>%
                                mutate(testing_level = case_when(
                                        between(test_per_positive, 0, 15) ~ "Below WHO recommendation",
                                        T ~ "On-par or Better than\n WHO recommendation"
                                ))

corr_result <- cor(correlation_test$test_per_positive, correlation_test$new_cases, use="complete.obs", method = "kendall")

correlation_test %>%
        filter(new_cases > 10, test_per_positive < 50, test_per_positive > 0) %>%
        ggplot() +
        geom_smooth(aes(x = test_per_positive, y = new_cases), size = 2) +
        geom_point(aes(x = test_per_positive, y = new_cases), size = 0.25) +
        theme_classic()+
        theme(legend.position = "none") +
        labs(x = "Test Per Positive", 
             y = "Number of daily cases(log10)",
             title = "Correlation between number of daily COVID-19 and test per positive",
             caption = "Data source: Our World in Data") +
        theme(plot.title = element_text(size = 12),
              text = element_text(family = "sans", color = "gray25"),
              plot.caption = element_text(color = "gray30"),
              plot.background = element_rect(fill = "gray95"),
              plot.margin = unit(c(5, 10, 5, 10), units = "mm")) +
        scale_y_log10()
        
```

Correlation co-efficent is good enough to indicate that something is going-on.  
Let's perform *hypothesis t-test* to confirm weather the correlation I recieved was sheer luck or not. 

```{r}
inference_result <- correlation_test %>% 
        filter(new_cases < 1000) %>% 
        inference(x = testing_level, y = new_cases, statistic = "mean", type = "ht", alternative = "greater", method = "theoretical", verbose =  F, null = 0)
```

The hypothesis t-test returned a p-value less than 0.001. This means that if tests per positve had no impact on  new_cases , then there less than 0.1% probability that we will observe this kind of data.

We can safely conclude that *India should aim for testing 15 or more people per positive case*.  

## India's current state

Let's have a look at the number of COVID-19 cases in Indian states and the level of testing.

```{r data_cleanup_india, echo=FALSE}
india_covid_data <- india_covid_data %>%
                        clean_names()
india_covid_data <- india_covid_data %>%
                        mutate(date = dmy(updated_on))
india_covid_data <- india_covid_data %>%
                        mutate(new_tests = total_tested - lag(total_tested)) %>%
                        mutate(new_cases = positive - lag(positive)) %>% 
                        mutate(days = round(as.numeric(difftime(date, "2020-04-17", unit="days")))) %>%
                        mutate(new_tests = ifelse(new_tests<0, 0, new_tests)) %>%
                        mutate(new_cases = ifelse(new_cases<0, 0, new_cases))

filtered_india_covid <- india_covid_data %>%
                                filter(new_cases != 0, new_tests != 0) %>%
                                mutate(test_per_positive = new_tests / new_cases)

latest_india_testing <- filtered_india_covid %>%
                                group_by(state) %>%
                                filter(days > max(days)-5) %>%
                                summarise(testing = mean(new_tests), 
                                          test_per_positive = mean(test_per_positive),
                                          new_cases = mean(new_cases))
latest_india_testing <- latest_india_testing %>%
                                mutate(tests_level = case_when(
                                        between(test_per_positive, 0, 15) ~ "Below WHO standard",
                                        between(test_per_positive, 16, 30) ~ "On-par with WHO standard",
                                        T ~ "Better than WHO standard"
                                )) %>%
                                mutate(tests_level = factor(tests_level, 
                                                              levels = c("Below WHO standard", 
                                                                         "On-par with WHO standard", 
                                                                         "Better than WHO standard")))
                        
latest_india_testing %>% 
                mutate(state = recode(state, "Dadra and Nagar Haveli and Daman and Diu" = "Daman and Diu",
                                      "Andaman and Nicobar Islands" = "Andaman and Nicobar")) %>%
                ggplot() +
                geom_col(aes(x = state, y = new_cases, fill = tests_level)) +
                theme_classic()+
                theme(axis.text.x = element_text(angle = 90,hjust=1, vjust = 0,size=10)) +
                labs(x = "States",
                     y = "Number of cases",
                     title = "Number of daily COVID-19 cases in different Indian states",
                     subtitle = "From 20 Jun to 25 Jun",
                     caption = "Data source: covid19india.org") +
                theme(plot.title = element_text(size = 12),
                      text = element_text(family = "sans", color = "gray25"),
                      plot.caption = element_text(color = "gray30"),
                      plot.background = element_rect(fill = "gray95"))+
                scale_fill_manual(values = c("red", "#FFE200", "#00C700"),
                          name = "Testing Level",
                          labels = c("Below WHO \nrecommendation(<15)",
                                     "On-par with WHO\n recommendation(15-30)",
                                     "Better than WHO\n recommendation(>30)"))

```

Clearly, all the states with high number of COVID-19 cases are struggling to get perform required amount of testing.

From the previous analysis, we established that performing 15 or more tests per positive helps in controlling the number of COVID-19. Therefore, assuming 15 to be bare miniumum of tests required to be performed per positive, let's have a look at how different states are performing.(To ease the understanding, the data of states that have less than 100 daily cases has been removed.)

```{r, echo=FALSE}
testing_required <- latest_india_testing %>%
                                mutate(testing_required = (testing/test_per_positive)*15) %>%
                                mutate(testing_difference = ((testing_required - testing)/testing))  
testing_required %>% 
                filter(new_cases > 100) %>%
                ggplot() +
                geom_col(aes(x = state, y = testing, color = "cyan"), fill = "cyan") +
                geom_col(aes(x = state, y = testing_required, color = "red"), fill = NA) +
                geom_text(aes(x = state, y = testing_required, 
                              label = scales::percent(testing_difference, accuracy = 1)), 
                        vjust = -1, size = 3)+
                theme_classic()+
                labs(x = "States",
                     y = "Number of tests",
                     title = "Amount of COVID-19 test deficit in different Indian states",
                     subtitle = "From 20 Jun to 25 Jun",
                     caption = "Data source: covid19india.org") +
                theme(plot.title = element_text(size = 12),
                      text = element_text(family = "sans", color = "gray25"),
                      plot.caption = element_text(color = "gray30"),
                      plot.background = element_rect(fill = "gray95"))+
                theme(axis.text.x = element_text(angle = 90,hjust=1, vjust = 0, size=10))+
                ylim(0,80000) +
                scale_color_identity(
                          breaks = c("red", "cyan"),
                          labels = c("Recommended Testing", "Current Testing"),
                          guide = "legend",
                          name = NULL)
```

States such as Delhi and Maharashtra are missing the required target level by more than 150%. States such as Andra Pradesh, Uttar Pradesh, and West Bengal have stepped to the task.

## Conclusion and Recommendation

To contain the spread of COVID-19 further in the month of July:  

1. India needs to increase it's testing capacity.  
2. Indian states can rethink their testing strategy as well. States should work as a single central body. They can pool in their resources to collectively reach their testing target. E.g. states like Andhra Pradesh can perform testing for Mumbai and Telanaga patients as well.  
3. India can use strategic lockdowns to increase it's testing capacity.